Circularity in Voxel-Based Analysis

### 2011-10-19 ANTsTBSS/paperRevision.tex
ENH:  MInor edits.

\begin{abstract}
Recent discussions within the neuroimaging community have highlighted the
problematic presence of selection bias in experimental design.  Although 
initially centering around the selection of voxels during the course of
fMRI studies, we demonstrate how this concept of circularity bias can 
potentially corrupt voxel-based analysis with a particular emphasis on
population studies of fractional anisotropy (FA).  
For such studies, template-based registration plays a critical role
in which a representative template serves as the normalized space for 
group alignment.  Following 
computation of FA, each subject's FA map is registered to the FA template prior to 
performing statistical comparisons between different groups within a population.
We illustrate how this analysis strategy can induce selection
bias which varies in strength with the similarity metric used in
registration.  Specifically, we analytically demonstrate that the popular sum of squared
difference (SSD) intensity metric explicitly maximizes effect size rather than
anatomical alignment---an experimental design flaw referred to as `circularity bias.'  
Through the use of both simulated diffusion data and real data involving
traumatic brain injury survivors, we explore the magnitude of this bias
and how it varies with similarity metric under the hypothesis that while
SSD-related metrics, such as Demons, will manifest similar effects, 
other metrics which 
are not formulated based on absolute intensity differences 
(e.g. mutual information (MI) and cross correlation (CC)) will produce
less of an effect. 
Consequently, given the variability in voxel-based analysis outcomes 
with similarity metric choice,
we caution researchers specifically in the use of SSD 
and SSD-related measures where normalization and statistical 
analysis involve the same image set and, instead, advocate a more 
cautious approach
where normalization of the individual subject images
to the statistical reference space occurs through the 
corresponding (statistically independent) anatomical images to 
an anatomical template.

@@

\begin{keyword}
Advanced Normalization Tools (ANTs) \sep circularity\ diffusion tensor imaging \sep fractional anisotropy \sep template construction \sep voxel-based analysis
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

The seminal work of Basser et al. \citep{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \citep{Basser1996,Assaf2008} enables promising analysis possibilities assessing neuro-structural differences in cross-population studies \citep{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010}.
% including subjects with traumatic brain injury (TBI) \citep{Caeyenberghs2010,Jiang2010,Warner2010}.  
Two popular comprehensive software packages for assessing population FA differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \citep{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM \citep{Ashburner2001} continues to find application to FA studies \citep{Kakeda2010,Takao2010,Preziosa2011} despite concerns about its general validity \citep{Bookstein2001,Davatzikos2004} and with respect to DTI \citep{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \citep{Arnone2008,Bodini2009}).

@@

However, suppose the SSD metric is used to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$, for statistical testing between groups. This results in the transformations,
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}}
\end{align}
is that which minimizes the voxelwise group variance.  Reducing the group variance will also directly affect the value of a standard population statistic, the Student's $t$-test.  Based on this derivation, one would also expect the popular Demons metric to be susceptible to this erroneous increase in statistical power due to identical objective functions although a more aggressive displacement/gradient at each iteration will relatively exacerbate this effect for a given number of iterations.  
Other standard similarity metrics will also exhibit this
trend to some degree as they minimize some local measure of intensity-based distance between the subject and the template although the effect will be minimized relative to the similarity to Eqn. (\ref{eq:variance}).

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct normalization of the images to be statistically compared 
explicitly conflates
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \citep{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  As illustrated with
Eqn. \ref{eq:variance}, direct FA-to-FA template registration will produce transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's 
$t$-test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are strikingly important in terms of the effect on detection power.

@@

Instead of direct FA-to-FA template mapping for 
alignment to the normalized space, we advocate a statistically 
independent alignment protocol in which the FA images are mapped to 
the common reference space via intra-subject FA-to-T1 mapping 
followed by T1-to-T1 template alignment thus eliminating all direct
FA-to-FA registrations. The key point is that the intensity value differences used 
to assess alignment (i.e. the similarity metric) in FA-to-T1 and T1-to-T1 
normalizations are statistic gally independent of the intensity value differences used for 
the post-alignment numerical analysis.  It should be noted that this discussion
is certainly not limited to studies involving FA.  It should be apparent
that such bias can occur anytime the images used for normalization
are non-independent of the statistical analysis as defined by the similarity
metric and statistical testing.

Two sets of experiments establish that the similarity metric
can significantly affect voxel-based analysis outcomes and that,
based on the previous description of circularity bias,
the relationship in performance variability and similarity metric
is to be expected.  Specifically, we show that whereas SSD and Demons
substantially overestimate statistical significance, mutual
information and cross correlation are less susceptible although the
effect is not
negligible.  Furthermore, we also show that this effects persists
even with the common practice of post-normalization smoothing prior
to mass univariate statistical testing.

Given this general overview, the experimental sets can be briefly
characterized as follows:
\begin{itemize}
  \item {\it Simulated DTI Data}:  Using a simulated DTI generator,
  FA images based on male/female populations were generated in situ
  thus guaranteeing anatomical alignment prior to processing.
  The simulated images in each group were then ``registered'' to the FA image
  of the defining DTI 
  template using four common similarity metrics (CC, Demons, MI, and SSD).  
  ROC curves based on voxel-based analysis for each of these scenarios demonstrate that 
  effect size is maximized over anatomical alignment and that
  this effect size varies with similarity metric.  Additional
  experiments establish that this trend persists even for commonly used smoothing
  kernels and the projection step of TBSS.
  \item {\it Real TBI Data}:  FA data of TBI survivors and controls were 
  normalized to a template using four common similarity metrics
  (CC, Demons, MI, and SSD).  Additional normalization included normalization via
  the subjects' anatomical T1-weighted images as well as the default registration for TBSS.
  Voxelwise statistical testing was also performed with
  and without smoothing.  Even though ground truth is not available for such
  data, the SSD and Demons metrics showed increased statistical power as evidenced
  by a general decrease in voxelwise $p$-values.
\end{itemize}
All major processing steps were performed using the open source Advanced Normalization Tools 
(ANTs)%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
software repository which should facilitate reproducibility for the interested 
reader.

@@

The TBI data used in this study is part of a larger effort investigating the relationship between various neuroimaging indices and cognitive and functional abilities in long-term survivors of TBI (principal investigator: John Whyte). A more detailed data description can be found in \cite{Avants2008}. The following is a brief description relevant to our particular study.  
17 controls and 16 patients with TBI were used for the analysis presented in this work.  Each patient had a history of non-penetrating traumatic brain injury 
of at least moderate severity defined by significant and well-documented
loss or alteration of consciousness following injury in addition to meeting
several other exclusionary criteria.  The healthy volunteers were matched in terms of age, gender, ethnicity, handedness, and years of education.
High resolution T1-weighted anatomic images were obtained using a 3D MP-RAGE 
imaging sequence with the following acquisition parameters: TR = 1620 ms, 
TI = 950 ms, TE = 3 ms, flip angle = 15$^\circ$, 160 contiguous slices of 
1.0 mm thickness, FOV = 192 $\times$ 256 mm$^2$, matrix = 192 $\times$ 256, 
1NEX with a scan time of 6 minutes, and voxel size = 1 mm$^3$.  30-directional
DTI images were also obtained. Diffusion tensor data were reconstructed using Camino.

@@

To provide realistic ground truth FA data, we implemented 
the method of Van Hecke et al. \cite{van-hecke2009} which 
we have also made available to the public
\cite{Tustison2011}.%
\footnote{
http://hdl.handle.net/10380/3315
}  
Given a DTI atlas, a b0 image, a scheme file indicating
the gradient directions, and 
a cohort of aligned DTI data (to model intersubject variability), the
algorithm generates
simulated data sets with specified pathology and noise.
For the simulated experiments described in the next section we
created a T1-weighted template from 30 randomly chosen T1-weighted images of
the NKI/Rockland data set described above.  
The resulting
template was then registered to the MNI152 template.  
The remaining 32 T1-weighted images were also registered to the
template resulting in a composite transform for each subject describing
the transform from the individual subject's T1-weighted image domain to 
the MNI152 template.

To obtain the mappings from the DTI space to the population-specific
template, the following steps were taken:
\begin{itemize}
  \item For each anatomical T1-weighted image, segment the white matter (in ANTs, one can use N4 bias correction \citep{Tustison2010} and Atropos $n$-tissue segmentation \citep{Avants2011a}).
  \item Rigidly transform each subject's average DWI to the bias-corrected anatomical image ($\mathcal{R}^i_{dwi}$).
  \item Perform distortion correction on the FA to anatomical mapping 
        by localized nonrigid registration
        of rigidly transformed FA map ($=\mathcal{R}^i_{dwi} \mathrm{FA}$) using
        the white matter mask obtained in step 3(a) ($\mathcal{T}^i_{FA}$).
  \item For each subject, concatenate the transforms which maps each
        subject's FA to the template ($\mathcal{T}^i_{total} = \mathcal{T}^i_{template} \circ 
        \mathcal{T}^i_{FA} \circ \mathcal{R}^i_{dwi} \mathrm{FA}$).
  \item Warp each subject's FA map to the template using $\mathcal{T}^i_{total}$.  
%  Equivalently, one could warp the mean diffusion (MD), axial diffusion (AD) or any 
%  other scalar derivable from DTI to the template.  
\end{itemize}
DTI 
were reoriented using the PPD method \cite{alexander2001}.  The 62
aligned DTI were averaged using the log-euclidean framework 
\cite{arsigny2006} to create the DTI atlas.  Similarly, the 62 aligned
b0 images were averaged to create the representative b0 image
for simulated DTI creation.  The derived FA image from the DTI atlas and 
the b0 image are shown in 
Fig. \ref{fig:simulated}.  An FA mask was created by thresholding 
the FA image of the DTI atlas at 0.2.

@@

\begin{figure*}
\begin{center}
\begin{tabular}{cc}
  \includegraphics[width=86mm]{ISV_SpeSenAlreadyAligned.pdf} &
  \includegraphics[width=86mm]{ISV_SpeSen_projectionAlreadyAligned.pdf}
\end{tabular}
\caption{ROC curves plotting $(1 - specificity)$ vs. $sensitivity$ for the
different similarity metrics for both (a) voxelwise analysis over the
entire FA mask and (b) using TBSS projections onto the mean FA skeleton.
Degree of locality and correspondence to 
Eqn. (\ref{eq:variance}) resulted in performance disparity between the
different metrics.  Since SSD and the SSD-like Demon's formulation 
minimize voxelwise group variance, we see the greatest bias in those 
two metrics.  In contrast, MI and CC are less susceptible to bias with
MI performing quite well.  Comparing voxelwise and projection results, 
it is apparent that the TBSS projection strategy only slightly improves 
the effects of bias for the simulated data.
}
\label{fig:simulated_plots}
\end{center}        
\end{figure*}

\begin{figure*}
\begin{center}
\begin{tabular}{ccc}
  \includegraphics[width=57mm]{ISV_SpeSenAlreadyAlignedSmooth0.pdf} &
  \includegraphics[width=57mm]{ISV_SpeSenAlreadyAlignedSmooth1.pdf} &
  \includegraphics[width=57mm]{ISV_SpeSenAlreadyAlignedSmooth2.pdf} \\
  \includegraphics[width=57mm]{ISV_CDFAlreadyAlignedSmooth0.pdf} &
  \includegraphics[width=57mm]{ISV_CDFAlreadyAlignedSmooth1.pdf} &
  \includegraphics[width=57mm]{ISV_CDFAlreadyAlignedSmooth2.pdf} 
\end{tabular}
\caption{Post-normalization smoothing analysis of the simulated data for
fwhm kernel sizes of 4 mm, 8 mm, and 16 mm.  Top row:  
ROC curves plotting $(1 - specificity)$ vs. $sensitivity$ for the
different similarity metrics for voxelwise analysis.  Smoothing
tends to ameliorate the bias effect but the relative strength
continues to correlate with similarity metric.  Bottom row:
A relative 
increase in (false) statistical power induced by SSD and Demons, as 
described in the text, resulted in more voxels having lower 
$p$-values.  This will be manifested as a left-shifted CDF 
curve which is apparent in the SSD and Demons metrics for all selected smoothing kernel sizes.
}
\label{fig:simulated_plots_smooth}
\end{center}        
\end{figure*}

\begin{figure}
\begin{center}
\begin{tabular}{cc}
  \includegraphics[width=41mm]{smooth0_truth.png} &
  \includegraphics[width=41mm]{smooth0_demons.png} \\
  (a) & (b)
\end{tabular}
\caption{(a) Mid-axial slice of the DTI atlas-derived FA map where the 
significant differences ($p \leq 0.05$, smoothing = 4 mm fwhm) between control (male) and 
experimental (female) groups are labeled in red.  (b) Same mid-axial 
slice showing an increase in significant regions following ANTs 
registration employing the Demons metric.
}
\label{fig:simulated_significance}
\end{center}        
\end{figure}

@@

Voxelwise analysis included performing Student's t-test within the
DTI masked region (FA $\geq 0.2$) for both the ground truth (i.e.
anatomically aligned) data set and performing the same calculations
for each of the four registration scenarios.  We used a significance
level cutoff of $p = 0.05$ to determine ground truth significant
voxels in the already aligned data set.  We then varied the 
significance level cutoff for each of the registered data sets to plot
the corresponding ROC curves given in Fig. \ref{fig:simulated_plots}.
Visual comparison of the significant regions associated with the 
voxelwise analysis for ground truth data and with the Demons registration data for
$p \leq 0.05$ shown in Fig. \ref{fig:simulated_significance} demonstrates 
the potential severity metric choice can have on results.

@@

Common practice is to smooth the images following normalization and
prior to statistical testing where the kernel size is typically dependent upon
one's estimate of the registration accuracy.  We tested the application
of common kernel sizes of 4 mm, 8 mm, and 16 mm full width at half
maximum (fwhm).  These plots are shown in Fig. \ref{fig:simulated_plots_smooth}
where we have produced similar ROC curves as in Fig. \ref{fig:simulated_plots}.
We also included $p$-value cumulative distribution functions (CDFs) \cite{yanovsky2009}.
Whereas ROC curves incorporate false/true positives and false/true
negatives, the $p$-value CDFs directly illustrate that certain metrics 
induce false positives.  False elevation of statistical power is caused
by a general decrease in $p$-values.  This decrease in $p$-values causes
a left-shift in the CDF which is observed for both the SSD and Demons metrics.

@@

For each of these scenarios, the normalized images were smoothed using
a kernel size of 3.5 mm (fwhm).  
Voxelwise analysis was then performed
within the region defined by thresholding the rescaled FMRIB58 template
(rescaled to $[0,1]$, thresholded at FA $\geq$ 0.2). Application of
 Student's t-test 
(one-tailed, $\alpha = 0.95$)
 at each voxel between the smoothed experimental and control group
 FA images
resulted in a set of $p$-values for each experiment.  Since standard
practice typically applies multiple comparisons correction to this
mass univariate statistical testing, we used
false discovery rate (FDR) correction to produce a set of corrected
$p$-values.

@@

The CDF of the voxelwise FA coefficient of variation (CV) of the aligned
images is shown for each case in Figure \ref{fig:tbi_cov}.  
The CV at each voxel is defined as the variance
divided by the mean over the population of FA values.  Thus,
those metrics which best minimize the voxelwise variance would
trend towards greater alignment as quantified by the CV \cite{van-hecke2011}.
Since minimization of the groupwise variance
will also minimize the CV, it is expected that the SSD and Demons 
metrics will produce relatively superior alignments to the
other strategies 
where alignment assessment is defined by the CV whereas the MI and CC metrics 
produce lower CV alignments.  Despite the fact that no direct FA-to-FA 
registrations were ever performed, the `antsT1' normalization produced better alignments
than `tbss'.  These can be visually assessed in Fig. \ref{fig:meanFA} where
we plotted the resulting mean FA images for both sets of normalizations.
Although direct FA-to-FA registration with common
similarity metrics gives ``better alignment'' such alignment is 
produced at a cost of significant type 1 errors.

>>>>

\begin{abstract}
Recent discussions within the neuroimaging community have highlighted the
problematic presence of selection bias in experimental design.  Although 
initially centering around the selection of voxels during the course of
fMRI studies, we demonstrate how this bias can 
potentially corrupt voxel-based analysis with a particular emphasis on
population studies of fractional anisotropy (FA).  
For such studies, template-based registration plays a critical role
in which a representative template serves as the normalized space for 
group alignment.  Following 
computation of FA, each subject's FA map is registered to the FA template prior to 
performing statistical comparisons between different groups.
We illustrate how this analysis strategy can induce selection
bias which varies in strength with the similarity metric used in
registration.  Specifically, we analytically demonstrate that the popular sum of squared
difference (SSD) intensity metric explicitly maximizes effect size rather than
anatomical alignment---an experimental design flaw referred to as `circularity bias.'  
Through the use of both simulated diffusion data and real data involving
traumatic brain injury survivors, we explore the magnitude of this bias
and how it varies with similarity metric under the hypothesis that while
SSD-related metrics, such as Demons, will manifest similar effects, 
other metrics which 
are not formulated based on absolute intensity differences 
(e.g. mutual information (MI) and cross correlation (CC)) will produce
less of an effect. 
Consequently, given the variability in voxel-based analysis outcomes 
with similarity metric choice,
we caution researchers specifically in the use of SSD 
and SSD-related measures where normalization and statistical 
analysis involve the same image set and, instead, advocate a more 
cautious approach
where normalization of the individual subject images
to the statistical reference space occurs through the 
corresponding (statistically independent) anatomical images.

@@

\begin{keyword}
Advanced Normalization Tools (ANTs) \sep circularity \sep diffusion tensor imaging \sep fractional anisotropy \sep template construction \sep voxel-based analysis
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

@@

The seminal work of Basser et al. \citep{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \citep{Basser1996,Assaf2008} enables promising analysis possibilities assessing neuro-structural differences in cross-population studies \citep{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010}.
% including subjects with traumatic brain injury (TBI) \citep{Caeyenberghs2010,Jiang2010,Warner2010}.  
Two popular comprehensive software packages for assessing population differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \citep{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM \citep{Ashburner2001} continues to find application to FA studies \citep{Kakeda2010,Takao2010,Preziosa2011} despite concerns about its general validity \citep{Bookstein2001,Davatzikos2004} and with respect to DTI \citep{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \citep{Arnone2008,Bodini2009}).

@@

However, suppose the SSD metric is used to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$, for statistical testing between groups. This results in the transformations,
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution,
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}},
\end{align}
is that which minimizes the voxelwise group variance.  Reducing the group variance will also directly affect the value of a standard population statistic, the Student's $t$-test.%
\footnote{
Based on this derivation, one would also expect the popular Demons metric to be susceptible to this erroneous increase in statistical power due to identical objective functions although a more aggressive displacement/gradient at each iteration will relatively exacerbate this effect for a given number of iterations.  
Other standard similarity metrics will also exhibit this
trend to some degree as they minimize a local measure of intensity-based distance between the subject and the template.
}

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct normalization of the images to be statistically compared using SSD
explicitly conflates
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \citep{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  Direct FA-to-FA template registration will produce transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's 
$t$-test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are strikingly important in terms of the effect on detection power.

@@

Instead of direct FA-to-FA template mapping for 
alignment to the normalized space, we advocate a statistically 
independent alignment protocol in which the FA images are mapped to 
the common reference space via intra-subject FA-to-T1 mapping 
followed by T1-to-T1 template alignment thus eliminating all direct
FA-to-FA registrations. The key point is that the intensity value differences used 
to assess alignment (i.e. the similarity metric) in FA-to-T1 and T1-to-T1 
normalizations are statistically independent of the intensity value differences used for 
the post-alignment numerical analysis.  It should be noted that this discussion
is certainly not limited to studies involving FA.  
Such bias can occur anytime the images used for normalization
are non-independent of the statistical analysis as defined by the similarity
metric and statistical testing.

Two sets of experiments establish that the similarity metric
can significantly affect voxel-based analysis outcomes and that,
based on the previous description of circularity bias,
the relationship in performance variability and similarity metric
is to be expected.  Specifically, we show that whereas SSD and Demons
substantially overestimate statistical significance, MI and CC are less susceptible although the
effect is not
negligible.  Furthermore, we also show that this effects persists
even with the common practice of post-normalization smoothing prior
to mass univariate statistical testing.

Given this general overview, the experimental sets can be briefly
characterized as follows:
\begin{itemize}
  \item {\it Simulated DTI Data}:  Using a simulated DTI generator,
  FA images based on male/female populations were generated in situ
  thus guaranteeing anatomical alignment prior to processing.
  The simulated images in each group were then ``registered'' to the FA image
  of the defining DTI 
  template using four common similarity metrics (CC, Demons, MI, and SSD).  
  ROC curves based on voxel-based analysis for each of these scenarios demonstrate that 
  effect size is maximized over anatomical alignment and that
  this effect size varies with similarity metric.  Additional
  experiments establish that this trend persists even for commonly used smoothing
  kernels and the TBSS projection step.
  \item {\it Real TBI Data}:  FA data of TBI survivors and controls were 
  normalized to a template using four common similarity metrics
  (CC, Demons, MI, and SSD).  Additional exploration included normalization via
  the subjects' anatomical T1-weighted images as well as the default registration for TBSS.
  Voxelwise statistical testing was also performed with
  and without smoothing.  Even though ground truth is not available for such
  data, the SSD and Demons metrics showed increased statistical power as evidenced
  by a general decrease in voxelwise $p$-values.
\end{itemize}
All major processing steps were performed using the open source Advanced Normalization Tools 
(ANTs)%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
software repository which should facilitate reproducibility for the interested 
reader.

@@

The TBI data used in this study is part of a larger effort investigating the relationship between various neuroimaging indices and cognitive and functional abilities in long-term survivors of TBI (principal investigator: John Whyte). 
17 controls and 16 patients with TBI were used for the analysis presented in this work.  Each patient had a history of non-penetrating traumatic brain injury 
of at least moderate severity defined by significant and well-documented
loss or alteration of consciousness following injury in addition to meeting
several other exclusionary criteria.  The healthy volunteers were matched in terms of age, gender, ethnicity, handedness, and years of education.
High resolution T1-weighted anatomic images were obtained using a 3D MP-RAGE 
imaging sequence with the following acquisition parameters: TR = 1620 ms, 
TI = 950 ms, TE = 3 ms, flip angle = 15$^\circ$, 160 contiguous slices of 
1.0 mm thickness, FOV = 192 $\times$ 256 mm$^2$, matrix = 192 $\times$ 256, 
1NEX with a scan time of 6 minutes, and voxel size = 1 mm$^3$.  30-directional
DTI images were also obtained. Diffusion tensor data were reconstructed using Camino.
A more detailed data description can be found in \cite{Avants2008}.

@@

To provide realistic ground truth FA data, we implemented 
the method of Van Hecke et al. \cite{van-hecke2009} which 
we have also made available to the public
\cite{Tustison2011}.%
\footnote{
http://hdl.handle.net/10380/3315
}  
Given a DTI atlas, a b0 image, a scheme file indicating
the gradient directions, and 
a cohort of aligned DTI data (to model intersubject variability), the
algorithm generates
simulated data sets with specified pathology and noise.
For the simulated experiments described in the next section we
created a T1-weighted population-specific template from 30 randomly chosen T1-weighted images of
the NKI/Rockland data set previously described.  
The resulting
template was then registered to the MNI152 template.  
The remaining 32 T1-weighted images were also registered to the population-specific 
template resulting in a composite transform for each subject describing
the mapping from the individual subject's T1-weighted image domain to 
the MNI152 template.

To obtain the mappings from the DTI space to the population-specific
template, the following steps were taken:
\begin{itemize}
  \item For each anatomical T1-weighted image, segment the white matter (in ANTs, one can use N4 bias correction \citep{Tustison2010} and Atropos $n$-tissue segmentation \citep{Avants2011a}).
  \item Rigidly transform each subject's average DWI to the bias-corrected anatomical image ($\mathcal{R}^i_{dwi}$).
  \item Perform distortion correction on the FA to anatomical mapping 
        by localized nonrigid registration
        of rigidly transformed FA map ($=\mathcal{R}^i_{dwi} \mathrm{FA}$) using
        the white matter mask obtained in a previous step ($\mathcal{T}^i_{FA}$).
  \item For each subject, concatenate the transforms which maps each
        subject's FA to the template ($\mathcal{T}^i_{total} = \mathcal{T}^i_{template} \circ 
        \mathcal{T}^i_{FA} \circ \mathcal{R}^i_{dwi} \mathrm{FA}$).
  \item Warp each subject's FA map to the template using $\mathcal{T}^i_{total}$.  
%  Equivalently, one could warp the mean diffusion (MD), axial diffusion (AD) or any 
%  other scalar derivable from DTI to the template.  
\end{itemize}
DTI 
were reoriented using the PPD method \cite{alexander2001}.  The 62
aligned DTI were averaged using the log-euclidean framework 
\cite{arsigny2006} to create the DTI atlas.  Similarly, the 62 aligned
b0 images were averaged to create the representative b0 image
for simulated DTI creation.  The derived FA image from the DTI atlas and 
the b0 image are shown in 
Fig. \ref{fig:simulated}.  An FA mask was created by thresholding 
the FA image of the DTI atlas at 0.2.

@@

\begin{figure*}
\begin{center}
\begin{tabular}{ccc}
  \includegraphics[width=57mm]{ISV_SpeSenAlreadyAligned.pdf} &
  \includegraphics[width=57mm]{ISV_SpeSenAlreadyAlignedSmooth0.pdf} &
  \includegraphics[width=57mm]{ISV_SpeSenAlreadyAlignedSmooth1.pdf} \\
  \includegraphics[width=57mm]{ISV_CDFAlreadyAligned.pdf} &
  \includegraphics[width=57mm]{ISV_CDFAlreadyAlignedSmooth0.pdf} &
  \includegraphics[width=57mm]{ISV_CDFAlreadyAlignedSmooth1.pdf} \\
\end{tabular}
\caption{Post-normalization smoothing analysis of the simulated data for
fwhm kernel sizes of 0 mm, 4 mm, and 8 mm.  Top row:  
ROC curves plotting $(1 - specificity)$ vs. $sensitivity$ for the
different similarity metrics for voxelwise analysis.  Smoothing
tends to ameliorate the bias effect but the relative strength
continues to correlate with similarity metric.  Bottom row:
A relative 
increase in (false) statistical power induced by SSD and Demons, as 
described in the text, resulted in more voxels having lower 
$p$-values.  This will be manifested as a left-shifted CDF 
curve which is apparent in the SSD and Demons metrics for all selected smoothing kernel sizes.
}
\label{fig:simulated_plots_smooth}
\end{center}        
\end{figure*}

\begin{figure}
\begin{center}
\begin{tabular}{cc}
  \includegraphics[width=41mm]{smooth0_truth.png} &
  \includegraphics[width=41mm]{smooth0_demons.png} \\
  (a) & (b)
\end{tabular}
\caption{(a) Mid-axial slice of the DTI atlas-derived FA map where the 
significant differences ($p \leq 0.05$, smoothing = 4 mm fwhm) between control (male) and 
experimental (female) groups are labeled in red.  (b) Same mid-axial 
slice showing an increase in significant regions following ANTs 
registration employing the Demons metric.
}
\label{fig:simulated_projection}
\end{center}        
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=81mm]{ISV_SpeSen_projectionAlreadyAligned.pdf}
\end{tabular}
\caption{ROC curves plotting $(1 - specificity)$ vs. $sensitivity$ for the
different similarity metrics using TBSS projections onto the mean FA skeleton.
Degree of locality and correspondence to 
Eqn. (\ref{eq:variance}) resulted in performance disparity between the
different metrics.  Since SSD and the SSD-like Demon's formulation 
minimize voxelwise group variance, we see the greatest bias in those 
two metrics.  In contrast, MI and CC are less susceptible to bias with
MI performing quite well.  Comparing voxelwise (cf Fig. \ref{fig:simulated_plots_smooth}) 
and projection results, 
it is apparent that the TBSS projection strategy only slightly improves 
the effects of bias for the simulated data.
}
\label{fig:simulated_plots}
\end{center}        
\end{figure}

@@

Voxelwise analysis included performing Student's $t$-test within the
DTI masked region (FA $\geq 0.2$) for both the ground truth (i.e.
anatomically aligned) data set and performing the same calculations
for each of the four registration scenarios.  We used a significance
level cutoff of $p = 0.05$ to determine ground truth significant
voxels in the already aligned data set.  We then varied the 
significance level cutoff for each of the registered data sets.
Visual comparison of the significant regions associated with the 
voxelwise analysis for ground truth data and with the Demons registration data for
$p \leq 0.05$ shown in Fig. \ref{fig:simulated_significance} demonstrates 
the potential severity metric choice can have on results.

Standard practice is to smooth the images following normalization and
prior to statistical testing where the kernel size is typically dependent upon
one's estimate of the registration accuracy.  We tested the application
of common kernel sizes of 4 mm and 8 mm full width at half
maximum (fwhm) and juxtaposed those results with no smoothing in
Fig. \ref{fig:simulated_plots_smooth}.  ROC curves are given
for each of the three cases.
We also included $p$-value cumulative distribution functions (CDFs) \cite{yanovsky2009}.
Whereas ROC curves incorporate false/true positives and false/true
negatives, the $p$-value CDFs directly illustrate that certain metrics 
induce false positives.  False elevation of statistical power is caused
by a general decrease in $p$-values.  This decrease in $p$-values causes
a left-shift in the CDF which is observed for both the SSD and Demons metrics.

@@

@@

For each of these scenarios, the normalized images were smoothed using
a kernel size of 3.5 mm (fwhm).  
Voxelwise analysis was then performed
within the region defined by thresholding the rescaled FMRIB58 template
(rescaled to $[0,1]$, thresholded at FA $\geq$ 0.2). Application of
 Student's $t$-test 
(one-tailed, $\alpha = 0.95$)
 at each voxel between the smoothed experimental and control group
 FA images
resulted in a set of $p$-values for each experiment.  Since standard
practice typically applies multiple comparisons correction to this
mass univariate statistical testing, we used
false discovery rate (FDR) correction to produce a set of corrected
$p$-values.

@@

The CDF of the voxelwise FA coefficient of variation (CV) of the aligned
images is shown for each case in Figure \ref{fig:tbi_cov}.  
The CV at each voxel is defined as the variance
divided by the mean over the population of FA values.  Thus,
those metrics which best minimize the voxelwise variance would
trend towards greater alignment as quantified by the CV \cite{van-hecke2011}.
Since minimization of the groupwise variance
will also minimize the CV, it is expected that the SSD and Demons 
metrics will produce relatively superior alignments to the
other strategies.  Despite the fact that no direct FA-to-FA 
registrations were performed, the `antsT1' normalization produced better alignments
than `tbss'.  These can be visually assessed in Fig. \ref{fig:meanFA} where
we plotted the resulting mean FA images for both sets of normalizations.
Although direct FA-to-FA registration with common
similarity metrics gives ``better alignment'' such alignment is 
produced at a cost of significant type 1 errors.
